# Querying a Database

**Contents**

1. [Querying a Database in Julia](#query_julia)
1. [Querying a Database in SAS](#query_sas)
1. [Querying a Database in R](#query_r)
1. [Querying a Database in Python](#query_python)

Once you are connected to a database, you can query the tables that are there or the ones you have written. To review writing tables, see the previous chapter on [writing to a database][Writing to a Database].

In the code blocks below, we assume you've already connected to your database and have named the connection as shown in the third chapter on [connecting to a database][Connecting to a Database].

In this chapter, we discuss the *logistics* of passing a query onto the database and reading the output using each language, rather than the basics of how to use SQL A brief introduction to SQL can be found in the [background chapter][Background]. If you're new to SQL and want more resources, here are some references we recommend:

- [SQL Keywords Reference by w3schools](https://www.w3schools.com/sql/sql_ref_keywords.asp): where there are multiple options (i.e. MySQL or SQL Server / Oracle / MS Access) make sure you follow instructions for SQL Server
- [SQL Tutorial by w3schools](https://www.w3schools.com/sql/)
- [SQL Server Basics by sqlservertutorial](https://www.sqlservertutorial.net/sql-server-basics/)

## Julia{#query_julia}

To query a database in Julia, we use the `ODBC.query` function. This takes in the connection `dsn` and a raw string for the query. This can be a one-line string, but we recommend using tripple quotes to pass in a multi-line string so you can use pretty SQL syntax.

```julia
res = ODBC.query(dsn,"""
  select cyl, drat 
    from mtcars 
    where disp > 200
""") 
```

This can also be done LINQ style, using `ODBC.Query`, with a capital Q.

```julia
res=@from i in ODBC.Query(dsn,"select * from mtcars") begin 
    @where i.disp > 200
    @select {Cylinder=i.cyl, Drat=i.drat}
    @collect table 
end 
```

To get the same result, you could also query the entire table and filter and select in Julia.

```julia
res = ODBC.query(dsn,"select * from mtcars") |> 
@filter( _.disp > 200) |>
@select( :cyl, :drat) |>
DataFrame
```

## SAS{#query_sas}

In SAS, we query using `proc sql`. The following code block references the connection libname created in [the connections chapter][#connect_sas].

```sas
proc sql;
  create table mtcars as
  select * from conn2.mtcars ;
  quit;
  
libname conn2 clear;
```

In this example, we don't use the libname. This works just the same, but the syntax is a bit more convoluted.

```sas
proc sql;
  connect to odbc as conn
  required="Driver={ODBC Driver 17 for SQL Server};Address=24.205.251.117;Database=NickDb;UID=&id;PWD=&pass";

  create table event as
  select * from connection to conn
  (select * from events) ;

  disconnect from conn;
  
quit;
```

## R{#query_r}

In R, you can choose to read a full table or query a subset with a sql query. Both options will return a `data.frame`.

**Read full table**
```{r eval = FALSE}
data <- dbReadTable(conn, name = "iris")
head(data)
```

**Query a table**
```{r eval = FALSE}
species <- dbGetQuery(
  conn, 
  'select distinct Species from iris'
)
species
```

## Python{#query_python}

### Standard Python
In standard python, we pass a `select` statement to `cursor.execute`. To fetch the results as a list of rows, we use `cursor.fetchall`. Each row in this output is of the type `pyodbc.Row`, and with this data type we can access values by the column names of the table using dot notation.

```python
cursor.execute("select * from iris")
iris_out = cursor.fetchall()

print(type(iris_out[1]))

first_sepal_length = iris_out[1].SepalLength
```

### Pandas
With pandas, we can use the `read_sql` function to query tables. This function returns the query results as a pandas DataFrame, which is frequently a lot easier to work with than the `pyodbc.Row` data type mentioned above. We recommend using pandas for this, and will continue to use pandas for the following chapters.

```python
import pandas as pd

iris_df_out = pd.read_sql("select * from iris", conn)
```
