# Querying a Database

Once you are connected to a database, you can query the tables that are there or the ones you have written. To review writing tables, see the previous chapter on [writing to a database](04-writing.Rmd).

In the code blocks below, we assume you've already connected to your database and have named the connection as shown in the third chapter on [connecting to a database](03-connections.Rmd).

In this chapter, we discuss the *logistics* of passing a query onto the database and reading the output using each language, rather than the basics of how to use sql. If you're new to sql, please review the query syntax for MS SQL Server. Here are some references we recommend:

- [SQL Keywords Reference by w3schools](https://www.w3schools.com/sql/sql_ref_keywords.asp): where there are multiple options (i.e. MySQL or SQL Server / Oracle / MS Access) make sure you follow instructions for SQL Server
- [SQL Tutorial by w3schools](https://www.w3schools.com/sql/)
- [SQL Server Basics by sqlservertutorial](https://www.sqlservertutorial.net/sql-server-basics/)

## Julia

**Connect to Database**

```{julia}
dsn = ODBC.DSN("Driver={ODBC Driver 17 for SQL Server};Address=24.205.251.117;Database=$database;UID=$user;PWD=$pass;")
```
**Query and filter using server-side sql commands**

Note, lower case "query" returns a dataframe, as opposed to uppercase "Query"

```{julia}
res=ODBC.query(dsn,"""

select cyl, drat 
from mtcars 
where disp > 200
      
""") 
```

### Query database to select all rows, then filter/select rows in julia

**LINQ style**
```{julia}
res=@from i in ODBC.Query(dsn,"select * from mtcars") begin 
    @where i.disp > 200
    @select {Cylinder=i.cyl, Drat=i.drat}
    @collect table 
end 
```
**dyplr/tidyverse style** 
```{julia}
res = ODBC.query(dsn,"select * from mtcars") |> 
@filter( _.disp > 200) |>
@select( :cyl, :drat) |>
DataFrame
```


## R
In R, you can choose to read a full table or query a subset with a sql query. Both options will return a `data.frame`.

**Read full table**
```{r eval = FALSE}
data <- dbReadTable(conn, name = "iris")
head(data)
```

**Query a table**
```{r eval = FALSE}
species <- dbGetQuery(
  conn, 
  'select distinct Species from iris'
)
species
```

## Python

### Standard Python
In standard python, we pass a `select` statement to `cursor.execute`. To fetch the results as a list of rows, we use `cursor.fetchall`. Each row in this output is of the type `pyodbc.Row`, and with this data type we can access values by the column names of the table using dot notation.

```python
cursor.execute("select * from iris")
iris_out = cursor.fetchall()

print(type(iris_out[1]))

first_sepal_length = iris_out[1].SepalLength
```

### Pandas
With pandas, we can use the `read_sql` function to query tables. This function returns the query results as a pandas DataFrame, which is frequently a lot easier to work with than the `pyodbc.Row` data type mentioned above. We recommend using pandas for this, and will continue to use pandas for the following chapters.

```python
import pandas as pd

iris_df_out = pd.read_sql("select * from iris", conn)
```

## SAS

**Query Using Libname**

Make the database connection.
```{sas}
libname conn2 odbc
required ="Driver={ODBC Driver 17 for SQL Server};Address=24.205.251.117;Database=NickDb;UID=&id;PWD=&pass";
```
Pull the rows that you need and disconnect from the database.
```{sas}
proc sql;
create table event as
select * from conn2.events ;
quit;
libname conn2 clear;
```
**Query without libname**

While this means a libname is unecessary, the syntax is much more convoluted
```{sas}
proc sql;
connect to odbc as conn
required="Driver={ODBC Driver 17 for SQL Server};Address=24.205.251.117;Database=NickDb;UID=&id;PWD=&pass";

create table event as
select * from connection to conn
(select * from events) ;

disconnect from conn;
quit;

```
