# Connecting to a Database

The first step to reading or writing to a database is the connection.

Note that with Julia, R, and Python, we can ask the user to input the address, database name, username, and password each time the code is run. Asking the user to input credentials is good practice for scenarios when you're sharing code, working on a shared computer, posting code publically on GitHub or elsewhere, or using sensitive data. However, if the files you're working with are private and you feel comfortable, you can replace these with the raw strings of your credentials (e.x. `database = "database name"`).

## Julia

```{f, eval=FALSE, highlight=FALSE}
using ODBC

# Get user credentials
address = begin 
    print("Enter Address: ")
    readline()
end 

database = begin 
    print("Enter Database Name: ")
    readline()
end 

user = begin 
    print("Enter Username: ")
    readline() 
end 

crypt = Base.getpass("Enter Password")
pass = read(crypt,String)

# setting up database  
dsn = ODBC.DSN("Driver={ODBC Driver 17 for SQL Server};Address=$address;Database=$database;UID=$user;PWD=$pass;")

# disconnect when you're done using this connection
ODBC.disconnect()
```

## SAS

When using SAS, we can connect through proc sql or using a libname. We would recommend the latter, as the connection can be referenced throughout your sql file, as opposed to within one proc sql statement.

**Proc Sql**
```sas
/*from proc sql directly*/
proc sql;
connect to odbc as conn
required="Driver={ODBC Driver 17 for SQL Server};Address=<address>;Database=<database name>;UID=<username>;PWD=<password>";

create table event as
select * from connection to conn
(select * from events) ;

/*disconnect when you're done using this connection*/
disconnect from conn;
quit;
```

**Libname**
```sas
/*by using a libname*/
libname conn2 odbc
required ="Driver={ODBC Driver 17 for SQL Server};Address=<address>;Database=<database name>;UID=<username>;PWD=<password>";

proc sql;
create table event as
select * from conn2.events ;

/*disconnect when you're done using this connection*/
disconnect from conn2;
quit;
```

## R

Using R, we connect to the database using the `odbc` and `DBI` packages.

```{r,eval=FALSE}
library(odbc)
library(DBI)

address = rstudioapi::showPrompt("Address","Address")
database = rstudioapi::showPrompt("Database name","Database name")
uid = rstudioapi::showPrompt("Database username", "Database username")
pwd = rstudioapi::askForPassword("Database password")

# open connection
conn <- dbConnect(
  odbc(),
  Driver = "ODBC Driver 17 for SQL Server",
  Server = address,
  Database = database,
  UID = uid,
  PWD = pwd
)

# disconnect when you're done using this connection
dbDisconnect(conn)
```

## Python 

The driver used in Python may depend on your setup. If you're working on a PC, downloading the driver and connection from Microsoft directly should allow you to reference the driver name: 

`driver = "{ODBC Driver 17 for SQL Server}"` 

If you're working on a Mac or Linux device, you cannot do this without further setup. It's easiest to pass in the path to the `dylib` file, which should be installed to `/usr/local/lib` by default. With this setup, you will need to reference the driver like this: 

`driver = "/usr/local/lib/libmsodbcsql.17.dylib"`

If copying and pasting the code below, switch which `driver =` line is commented depending on your operating system.

```python
import pyodbc

driver = "{ODBC Driver 17 for SQL Server}" # PC Users
# driver = "/usr/local/lib/libmsodbcsql.17.dylib" # Mac Users

address = input("Address: ")
database = input("Database name: ")
username = input("Username: ")
password = getpass.getpass(prompt = "Password: ")

# open connection
conn = pyodbc.connect(
    ";".join([
        "Driver="+driver,
        "Address="+address,
        "Database="+database,
        "UID="+username,
        "PWD="+password
    ])
)

# get cursor
cursor = conn.cursor()

# disconnect when you're done using this connection
conn.close()
```


## JSL
```{r,eval=FALSE,highlight=FALSE}
opendatabaseconnection(
        "Driver={ODBC Driver 17 for SQL Server};Address=<address>;Database=<database>;UID=<username>;PWD=<password>"
)
```

